Dissociation between vertebrae number and body size in Australian blindsnakes

### 2024-08-22 manuscript.Rmd
incorporated Emma's suggestions and inputs

* Axial evolution an important aspect of the vertebrate body plan evolution – facilitates diversity and influences locomotion 
  <!-- Note the papers  that are about axial number relating to locomotion -->
in mammals [@galisFast2014 ], lacertids [@vandammeSpeed2002], gobies [@yamadaAdaptive2009;@forkerBendy2021], and chameloens [@molnarMorphological2023].

* Need a reference for mammals? Limited variation of a common theme that is constrained [@asherVariability2011;@naritaEvolution2005]
@liDivergent2023 found vertebral counts to differ significantly between major groups of mammals and within groups. 
* Reptiles, on the other hand, are more varied [@headEvolution2015]

* Elongate animals have special properties of vertebral number known as pleomerism, where body size and number of segments in vertebrae are correlated [@lindseyPleomerism1975]. 
Pleomerism has been observed in eels and fish [@wardEvolution2007;@wardDifferential2014;@reeceEvolutionary2013], in squamates [@pollyTesting2001;@headDissociation2007;@headEvolution2015;@bergmannVertebral2012;@lindellEvolution1994;@cernanskyVertebral2019], in salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015].

* Snakes are a good model to study axial evolution because they have a high number of vertebrae and a wide range of body size and shape.
* In elapid snakes [@sherrattVertebral2019;@sherrattMacroevolution2022;@hamptonInterspecific2019;@hamptonRegionalization2023]; colubrid snakes [@nash-hahnModification2024].

***[Specific background]***

Scolecophidians, often referred to in English as "blindsnakes", are characterised by their fossorial lifestyle, poor visual system, and highly specialised phenotypes adapted for a strictly fossorial existence [@mirallesMolecular2018;@strongDeconstructing2021;@gowerEyetranscriptome2021]. Scolecophidians comprise five paraphyletic families, including the "core scolecophidians" — a monophyletic clade consisting of Typhlopidae (425 spp), Leptotyphlopidae (143 spp), Gerrhopilidae (29 spp), and Xenotyphlopidae (1 spp) — and the paraphyletic Anomalepididae (20 spp), which is sister to all other "typical snakes" (Alethinophidia) [@mirallesMolecular2018; @burbrinkInterrogating2020; @uetzQuarter2021].

Although all blindsnakes are fossorial, body size and shape vary significantly among species [@shineNatural1990;@webbPreysize1993;@webbDietary2001;@ehmannFamily1993]. Leptophlopids are generally thought to be more narrow-and-long compared to typhlopids, which are generally wide-and-robust for a given body length [@herrelBurrowing2021]. These differences in body size and proportions determine how much push force the snakes can generate when burrowing. Their morphological differences led to the idea that leptyphlopids generally use existing burrows while typhlopids create their own burrow.

***[Knowledge gap]***

The Australian blindsnake radiation comprise 47 described species [@tiatragulAustralia2023] with considerably diverse morphology from narrow-and-long (aspect ratio >= 100) to robust-and-wide (aspect ratio < 60). However, information about the relationship between body size and vertebrae number is unknown.

Among typhlopidae, gigantism have evolved multiple times but with different mechanisms [@headDissociation2007]. For instance, @headDissociation2007 found that some of the largest species of blindsnakes in the African continent (e.g. *Afrotyphlops schlegelii*) have fewer than expected precloacal vertebrae than expected for their size. This suggests that elongation in these species is achieved by elongating each vertebra rather than adding additional vertebrae.

In another predominantly burrowing clade, caecilians, the number of vertebrae is positively correlated with body size [@renousBody1989], suggesting pleomerism is a mechanism for burrowing taxa to diversity in the underground environment.

<!-- A big review included number of vertebrae [@szyndlarIllustrated2023] for many non-alethinophidian snakes, but  -->

Compared to other snakes, the vertebral column of fossorial snakes tend to feature a much higher proportion of pre- than post-cloacal segments [@pollyTesting2001]. This is reflected in the relatively short tails of blindsnakes (CITE).

***[What we do in this paper]***

In this study, we investigate the prevalence of pleomerism among Australian blindsnakes and test hypotheses that deviations from pleomerism may be explained by different body shapes or other ecological factors among Australian typhlopid snakes. The specific questions we addressed are: (1) How prevalent is pleomerism among Australian blindsnakes? (2) Can deviations from pleomerism be explained by body shape in a genus of Australian blindsnakes? (3) Does vertebral ratio vary with soil compactness or mean annual temperature?

@@

To obtain vertebral counts, we X-ray scanned the specimens at various institutions including: LX-60 (Faxitron, IL, USA) at the University of Adelaide Faculty of Health and Medical Sciences (n=87), Thermo Kevex PXS5-927EA Microfocus source (Waltham, MA, USA) with a LTX-1717 Digital Flat Panel Detector (Kodex Inc, NJ, USA) at the Western Australian Museum (n=129), HeliScan MicroCT system at the National Laboratory for X-Ray Micro Computed Tomography (CTLab) based at the Australian National University (ANU) (n=90), and Queensland X-Ray (n=35).

@@

Additionally, we quantified intracolumnar length of each vertebra for three species representing 1) narrow-and-wide phenotypes (*A. grypus*), 2) wide-and-robust phenotypes (*A. pinguis*), and 3) intermediate phenotypes (*A. centralis*). Following @sherrattMacroevolution2022 we used the 'multipoint tool' in ImageJ2 to place landmarks medially along the vertebral column at the anterior limit of the centrum of each vertebra. We then calculated the inter-landmark distances by applying the Pythagorean theorem between sequential landmark coordinates in R Statistical Environment v.4.3.1 [@rcoreteamLanguage2023]. Measurements were repeated three times to account for measurement errors.

## Environmental data 
<!-- NEED TO PUT IN THE CORRECT VALUES HERE -->

@@

<!-- Fig. Vertebral counts vs max body length  -->
```{r fig-tbl-vert, fig.cap= 'Maximum vertebral counts plotted against maximum total body length for Australian typhlopids. Trend line represent PGLS regression for Australian species.', out.width="165mm", out.height= "124mm"}

@@

<!-- Fig: Pleomerism and ratio correlations -->
```{r fig-vert-aspect-ratios, fig.cap= 'Species mean aspect ratios against (A) mean vertebrae ratio and (B) mean number of vertebrae of Australian blindsnakes. Vertebrae ratio is calculated by total vertebrae/total length, and aspect ratio is total body length/midbody diameter. (C) Vertebral profile shape of three species representing different vertebrae ratios, and are depicted as a 4° polynomial curve with positive inflection. Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and-wide.', out.width="124mm", out.height= "165mm"}

@@

Body shape varied with the number of vertebrae in similar ways as previously reported in vipers and colubrids [@lindellEvolution1994]. 
We can speculate that species that are narrow-and-long may be using pre-existing burrows rather than digging their own burrow. Narrow-and-wide phenotypes are common among Leptotyphlopidae, which are thought to use existing burrows rather than dig their own burrow [@herrelBurrowing2021]. In this regard, the Australian typhlopid genus, *Anilios*, is quite exceptional because there is a great variation in body shape (Tiatragul et al. unpublished). Species that evolved more narrow-and-long bodies may be converging on the lifestyle with other scolecophidians that have similar phenotypes (i.e. leptotyphlopids). On the other hand, robust-and-wide phenotypes common among typhlopid species are expected to be able to generate much more force and make their own burrows [@herrelBurrowing2021].

Discuss amphisbaenians [@gansAmphisbaenia1978] and caecillians [@renousBody1989].

@@

Both mean annual temperature and soil bulk density were significant predictors of vertebrae ratio but in different directions. Mean annual temperature has been shown to be a proxy for several ecological factors such as rainfall and vegetation (CITE). However, temperature has also been shown to directly influence multiple axes of reptile development (cite). Vertebral development (somitogenesis) and homeotic (*Hox gene*) regulation may thus be affected by variations in incubation conditions [@martin-del-campoHox2018]. However, we currently lack knowledge about egg-laying behaviour and how temperature affects the development of blindsnake embryos. Common garden experiments across different species of blindsnakes may be able to shed light on whether variation in incubation temperature and consequently, the rate of development may alter the number of vertebrae formation during development. 
<!-- We would expect higher temperature to accelerate development (CITE anolis papers, turtles, etc), so less time for embryonic development.  -->

@@

## Miscellaneous 
Traditionally, ventral scale counts have been used as a proxy for vertebral number as they correlate for most alethinophidian snakes [@alexanderPattern1966]. We found that this is (not) the case in blindsnakes. Blindsnakes are generally small enough to fit in an X-ray scanner, even in portable field X-ray machines. Collecting more data on this group of basal snakes can help us understand how vertebral morphology evolved in snakes.

>>>>

Axial evolution is fundamental to vertebrate body plan evolution, facilitating body form changes and influencing modes of locomotion in mammals [@eg1500;@taewcharoenSmall2024;@galisFast2014]. reptiles [@eg1500; @vandammeSpeed2002;@molnarMorphological2023], amphibians [@eg1500;@bonettEvidence2017;@adlerEvolutionary2022] and fish [e.g., @yamadaAdaptive2009;@forkerBendy2021]. Variation of the axial skeleton is often studied by counting the number of vertebrae because the developmental mechanisms of segmentation are well characterised. Essentially vertebrae begin as somites, epithelial blocks of mesoderm, which form rhythmically during embryongenesis where the speed of this process relates to the size and number of somites, to produce what will become the axial skeleton [@gomezControl2008;@dequeantSegmental2008;@gomezDevelopmental2009]
Elongate animals are expected to show pleomerism, where body size and number of vertebrae are positively correlated [@lindseyPleomerism1975]. \
Pleomerism has been observed in eels and fish [@wardEvolution2007;@wardDifferential2014;@reeceEvolutionary2013], in squamates [@pollyTesting2001;@headDissociation2007;@headEvolution2015;@bergmannVertebral2012;@lindellEvolution1994;@cernanskyVertebral2019], in salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015]. However, there are notable exceptions in some taxa [@headDissociation2007;@sherrattMacroevolution2022] indicating it is possible to disassociate segment production with somatic (body) growth, and revealing developmental processes underlying evolutionary patterns. Snakes are therefore a good model to study axial evolution because they exhibit great diversity in vertebrae number across species and have a wide range of body sizes and shapes without limb skeleton to confound signals of locomotor adaptation. Recent studies have indicated the number of vertebrae varies predictably with ecological diversification in elapid snakes [@sherrattHeterochronic2019;@hamptonInterspecific2019;@sherrattMacroevolution2022] and colubrid snakes [@nash-hahnModification2024]. It also has important influence on snake locomotion, where the number of vertebrae negatively correlates with speed travelled in garter snakes [@kelleyEffects1997]. More research is required to understand the role that axial evolution plays in body plan and ecological diversification of snakes.

<!-- [Specific background]-->
Scolecophidians, commonly known in English as "blindsnakes" and "threadsnakes", are characterised by a reduced visual system and highly specialised phenotypes adapted for a strictly fossorial existence [@mirallesMolecular2018;@strongDeconstructing2021;@gowerEyetranscriptome2021]. Scolecophidians comprise five paraphyletic families, including the "core scolecophidians" — a monophyletic clade consisting of Typhlopidae (425 spp), Leptotyphlopidae (143 spp), Gerrhopilidae (29 spp), and Xenotyphlopidae (1 spp) — and the paraphyletic Anomalepididae (20 spp), which is sister to all other "typical snakes" (Alethinophidia) [@mirallesMolecular2018; @burbrinkInterrogating2020; @uetzQuarter2021]. 
Although all blindsnakes are fossorial, body size and shape vary significantly among species [Tiatragul et al. unpublished; @shineNatural1990;@webbPreysize1993;@webbDietary2001;@ehmannFamily1993]. Leptotyphlopids are generally thought to be more narrow-and-long compared to typhlopids, which are generally wide-and-robust for a given body length [@herrelBurrowing2021]. These differences in body size and proportions determine how much push force the snakes can generate when burrowing. Their morphological differences led to the idea that leptyphlopids generally use existing burrows while typhlopids create their own burrow.

In another predominantly burrowing clade, caecilians, the number of vertebrae is positively correlated with body size [@renousBody1989], suggesting pleomerism is a mechanism for burrowing taxa to diversity in the underground environment. 
Pleomerism has been observed in typhlopidae, however, some notable exceptions indicate different mechanisms for gigantism [@headDissociation2007]. For instance, Head and Polly [-@headDissociation2007] found that some of the largest species of blindsnakes in the African continent (e.g. *Afrotyphlops schlegelii*) have fewer than expected precloacal vertebrae than expected for their size. This suggests that elongation in these species is achieved by elongating each vertebra rather than adding additional vertebrae. Compared to other snakes, the vertebral column of fossorial snakes tends to feature a much higher proportion of pre- than post-cloacal segments [@pollyTesting2001]. This is reflected in the relatively short tails of blindsnakes [Tiatragul unpublished; @szyndlarIllustrated2023]. Limited studies on scolecophidian axial skeletons compared to alethinophidia means we have yet to appreciate the diversity that these groups have to offer. Given the recent shift in the perception of these earliest diverging clades of snakes (scolecophidians) away from the traditional view that they are "pleisomorphic and morphologically homogenous" [@strongConvergence2022], the scolecophidian axial skeleton also could offer a new perspective on the ecological origin of snakes.

The Australian blindsnake radiation comprise 47 described species [@tiatragulAustralia2023] with considerably diverse morphology from narrow-and-long (aspect ratio >= 100) to robust-and-wide (aspect ratio < 60). However, information about the relationship between body size and vertebrae number is unknown. In this study, we investigate the prevalence of pleomerism among Australian blindsnakes and test hypotheses that deviations from pleomerism may be explained by different body shapes or other ecological factors among Australian typhlopid snakes. The specific questions we addressed are: (1) How prevalent is pleomerism among Australian blindsnakes? (2) Can deviations from pleomerism be explained by body shape in a genus of Australian blindsnakes? (3) Does vertebral ratio vary with soil compactness or mean annual temperature?

@@

To obtain vertebral counts, we X-ray scanned the specimens at various institutions including: LX-60 (Faxitron, IL, USA) at the University of Adelaide Faculty of Health and Medical Sciences (n=87), Thermo Kevex PXS5-927EA Microfocus source (Waltham, MA, USA) with a LTX-1717 Digital Flat Panel Detector (Kodex Inc, NJ, USA) at the Western Australian Museum (n=129), HeliScan MicroCT system at the National Laboratory for X-Ray Micro Computed Tomography (CTLab) based at the Australian National University (n=90), and Queensland X-Ray (n=35).

@@

Additionally, we quantified intracolumnar length of each vertebra for three species representing (1) narrow-and-wide phenotypes (*A. grypus*), (2) wide-and-robust phenotypes (*A. pinguis*), and (3) intermediate phenotypes (*A. centralis*). Following @sherrattMacroevolution2022 we used the 'multipoint tool' in ImageJ2 to place landmarks medially along the vertebral column at the anterior limit of the centrum of each vertebra. We then calculated the inter-landmark distances by applying the Pythagorean theorem between sequential landmark coordinates in R Statistical Environment v.4.3.1 [@rcoreteamLanguage2023]. Measurements were repeated three times to account for measurement errors.

## Environmental data

@@

<!-- Fig. Vertebral counts vs max body length  -->
```{r fig-tbl-vert, fig.cap= 'Maximum vertebral counts plotted against maximum total body length for Australian typhlopids. Trend line represent PGLS regression for Australian species. Phylogeny projected into space using maximum likelihood ancestral state estimation. Species not included in the phylogeny are shown in purple. ', out.width="165mm", out.height= "124mm"}

@@

<!-- Fig: Pleomerism and ratio correlations -->
```{r fig-vert-aspect-ratios, fig.cap= 'Species mean aspect ratios against (A) mean vertebrae ratio and (B) mean number of vertebrae of Australian blindsnakes. Vertebrae ratio is calculated by total vertebrae/total length, and aspect ratio is total body length/midbody diameter. (C) Vertebral profile shape of three species representing different vertebrae ratios, and are depicted as a 4° polynomial curve with positive inflection. Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and-wide. Phylogeny projected into space using maximum likelihood ancestral state estimation. Species not included in the phylogeny are shown in purple.', out.width="124mm", out.height= "165mm"}

@@

Body shape varied with the number of vertebrae in similar ways as previously reported in vipers and colubrids [@lindellEvolution1994]. 
We speculate that species that are narrow-and-long may be using pre-existing burrows rather than digging their own burrow. Narrow-and-wide phenotypes are common among Leptotyphlopidae, which are thought to use existing burrows rather than dig their own burrow [@herrelBurrowing2021]. In this regard, the Australian typhlopid genus, *Anilios*, is quite exceptional because there is a great variation in body shape (Tiatragul et al. unpublished). Species that evolved more narrow-and-long bodies may be converging on the lifestyle with other scolecophidians that have similar phenotypes (i.e., leptotyphlopids). On the other hand, robust-and-wide phenotypes common among typhlopid species are expected to be able to generate much more force and make their own burrows [@herrelBurrowing2021].

Discuss amphisbaenians [@gansStudies1960] and caecillians [@renousBody1989].

@@

Both mean annual temperature and soil bulk density were significant predictors of vertebrae ratio but in different directions. Mean annual temperature has been shown to be a good proxy for ecological factors relevant to squamates [@coxEvolutionary2015;@pieEvolution2017;@wiensDiversity2013]. However, temperature has also been shown to directly influence multiple axes of reptile development [@nobleDevelopmental2018;@boothInfluence2006;@feinerDevelopmental2018]. Vertebral development (somitogenesis) and homeotic (*Hox* gene) regulation may thus be affected by variations in incubation conditions [@martin-del-campoHox2018]. However, we currently lack knowledge about egg-laying behaviour and how temperature affects the development of blindsnake embryos. Common garden experiments across different species of blindsnakes may be able to shed light on whether variation in incubation temperature and consequently, the rate of development may alter the number of vertebrae formation during development. 
<!-- We would expect higher temperature to accelerate development (CITE anolis papers, turtles, etc), so less time for embryonic development.  -->

@@

## Miscellaneous 
Traditionally, ventral scale counts have been used as a proxy for vertebral number as they correlate for most alethinophidian snakes [@alexanderPattern1966]. We found that this is not the case in blindsnakes.

Blindsnakes are generally small enough to fit in an X-ray scanner, even in portable field X-ray machines. Collecting more data on this group of basal snakes can help us understand how vertebral morphology evolved in snakes.
